---
title: "Statistics_adjuvant"
output: html_document
date: "2026-05-03"
---

# Statistical analysis
This code is for the statistical analysis that was performed on "Spatial Multiomic Profiling Identifies Metabolic and Inflammatory Signatures Driving Recurrence in Adjuvant-Treated NSCLC"
by Berrell, Tan, Killgallon et al. 2026. 

Specifically, this code refers to the statistical approach employed for the proteomic data analysis and uses mixed-effects modelling to account for instances of multiple samples per patient (pseudo-replicates). As these replicates were samples up to several centremeters apart, we opted for a mixed-effects model rather than averaging the scores to not dilute the within-patient biological heterogeneity. 

```{r setup, include=FALSE}

library(dplyr)
library(readxl)
library(ggplot2)
library(lme4)
library(lmerTest)
library(survival)
library(ClusterBootstrap)
library(survival)

# read the features - these have already been filtered by variance criteria, where over 60 values of each feature must be unique. 

features <- read_excel("features_filtered_updated_160226.xlsx")

#remove samples that are not Adenocarcinoma or Squamous
samples_to_remove <- c("Adenosquamous", "Large cell")

# Exclude these players from the dataframe
feature_full <- features[!(features$Histology %in% samples_to_remove), ]

```


## Preparing data

- Clean up the data here by fixing column names to not have any symbols. 
- create stage broad column: stage 1 and stage 2 = early, stage 3 = advanced
- filter the dataframe to exclude some of the metadata. Only to include time to relapse, relapse status, patient ID, Histology and stage

```{r}
#Cleaning data 
colnames(feature_full) <- gsub("@", "", colnames(feature_full), fixed=TRUE)
colnames(feature_full) <- gsub("+", "", colnames(feature_full), fixed=TRUE)
colnames(feature_full) <- gsub("~", "_to_", colnames(feature_full), fixed=TRUE)
colnames(feature_full) <- gsub(":", "in", colnames(feature_full), fixed=TRUE)
colnames(feature_full) <- gsub("&", "_", colnames(feature_full), fixed=TRUE)


feature_full$stage_broad <- ifelse(feature_full$stage_grouped =="stage3", "Advanced", "Early")

results_list_full_stage_hist <- list()


#Columns with numeric values-> filter to only these (should be features only and not clinical data)
numeric_cols <- sapply(feature_full, is.numeric)

df_numeric <- feature_full[, numeric_cols]

feature_vars <- setdiff(
  colnames(df_numeric),
  c("Relapse_followup...6922","Relapse_status...6921")
)


df_filtered <- feature_full[, c(
  "Relapse_followup...6922",
  "Relapse_status...6921",
  "patientID",
  "Histology",
  "stage_broad",
  "stage_grouped",
  feature_vars
)]

#scale the features only (not the clinical data)
df_filtered[feature_vars] <- scale(df_filtered[feature_vars])

```


# Cox model
Mixed-effects coxmodel assessing all samples + histology +stage 
- this model uses the cluster argument to account for multiple samples/patient
- code runs a loop for all features then applies bonferroni and Benjamini Hochberg p-value adjustments

```{r}

for (feature in feature_vars){
  formula_text <- paste0(
    "Surv(Relapse_followup...6922, Relapse_status...6921) ~ ",
        feature,
    " + Histology +stage_broad + cluster(patientID)"
  )
 model_full <- tryCatch(
    coxph(as.formula(formula_text), data = df_filtered),
    error = function(e) return(NULL)
  )
  
  if (is.null(model_full)) next
  
  summary_model_full <- summary(model_full)
  
  # Make sure the feature exists in the model output
  if (!(feature %in% rownames(summary_model_full$coefficients))) next
  
  coef_row <- summary_model_full$coefficients[feature, ]
  
  results_list_full_stage_hist[[feature]] <- data.frame(
    variable = feature,
    estimate = coef_row["coef"],
    std.error = coef_row["robust se"],   # IMPORTANT when using cluster()
    z = coef_row["z"],
    p.value = coef_row["Pr(>|z|)"],
    HR = exp(coef_row["coef"]),
    lower_CI = exp(coef_row["coef"] - 1.96 * coef_row["robust se"]),
    upper_CI = exp(coef_row["coef"] + 1.96 * coef_row["robust se"]),
    c_index <- summary_model_full$concordance[1]
  )
}

results_full_stage_hist <- do.call(rbind, results_list_full_stage_hist)


results_full_stage_hist$p_adjusted <- p.adjust(results_full_stage_hist$p.value, method = "bonferroni")

significant_findings_uni <- results_full_stage_hist %>% filter(p_adjusted<0.05)

results_full_stage_hist$fdrs<-p.adjust(results_full_stage_hist$p.value, method="BH")
write.csv(results_full_stage_hist, "stats_out/cox_allhistology.csv")

```



### Forest plot to display the results 
Plotting significant findings 

```{r plotting }
library(ggplot2)
library(dplyr)

# ── 1. Filter to significant hits ─────────────────────────────────────────────
sig_cox <- results_full_stage_hist %>% 
  filter(fdrs < 0.05)

# ── 2. Order variables by HR for cleaner plot ─────────────────────────────────
sig_cox <- sig_cox %>%
  mutate(variable = reorder(variable, HR))

# ── 3. Forest plot ────────────────────────────────────────────────────────────
p2 <- ggplot(significant_findings_uni, aes(x = HR, y = variable)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = lower_CI, xmax = upper_CI),
                 height = 0.25, linewidth = 0.7) +
  geom_point(aes(size = -log10(p_adjusted),
                 colour = ifelse(HR > 1, "Risk", "Protective"))) +
  scale_colour_manual(values = c("Risk" = "#d73027", "Protective" ="#4575b4"),
                      name = "Direction") +
  scale_x_log10() +
  scale_size_continuous(name = "-log10(p_adjusted)", range = c(2, 6)) +
  labs(
    x     = "Hazard Ratio (95% CI, log scale)",
    y     = NULL,
    title = "Significant Cox PH Results",
    subtitle = "Adjusted for Histology + Stage | FDR < 0.05"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.y = element_text(size = 7) 
  )

p2

```
### Condcordance plot 
This plot looks at the c index of the model compared with the adjusted p-values. Color of dots represent hazard ratios

- change colour to better show HRs around 0 

```{r}

p_condcordance <- ggplot(results_full_stage_hist, aes(x=c_index....summary_model_full.concordance.1., y=-log(fdrs)))+
  geom_point(aes(colour = log(HR)))+
  geom_hline(yintercept = -log(0.05), linetype = "dashed")+
  geom_vline(xintercept = 0.65)

p_condcordance

```
### Volcano plot 
Volcano plot that presents the log-fold adjusted p-values with their hazard ratios
- shapes indicate the type of feature 
```{r}
results_full_stage_hist$prefix <- substr(results_full_stage_hist$variable, 1, 4)


p_volcano <- ggplot(results_full_stage_hist, aes(x=log(HR), y=-log(fdrs), shape=prefix, label=variable))+
  geom_point()+ 
  #geom_text(nudge_x = 0.2, nudge_y = 0.2)+
  #geom_text_repel(max.overlaps = 70)+
  geom_hline(yintercept = -log(0.05), linetype = "dashed")
 
  #+
  #geom_vline(xintercept = 0.65)

p_volcano

```
# Performing cox calculations on adeno samples   
## All samples + adeno +stage 
```{r}
results_list_full_adeno <- list()

adeno <- df_filtered[df_filtered$Histology == 'Adenocarcinoma',]

numeric_cols_adeno <- sapply(adeno, is.numeric)

df_numeric_adeno <- adeno[, numeric_cols_adeno]

feature_vars_adeno <- setdiff(
  colnames(df_numeric_adeno),
  c("Relapse_followup...6922","Relapse_status...6921")
)


ad_df_filtered <- adeno[, c(
  "Relapse_followup...6922",
  "Relapse_status...6921",
  "patientID",
  "stage_broad",
  feature_vars_adeno
)]


ad_df_filtered[feature_vars_adeno] <- scale(ad_df_filtered[feature_vars_adeno])


for (feature in feature_vars_adeno){
  formula_text <- paste0(
    "Surv(Relapse_followup...6922, Relapse_status...6921) ~ ",
        feature,
    "+stage_broad + cluster(patientID)"
  )
 model_full <- tryCatch(
    coxph(as.formula(formula_text), data = ad_df_filtered),
    error = function(e) return(NULL)
  )
  
  if (is.null(model_full)) next
  
  summary_model_full <- summary(model_full)
  
  # Make sure the feature exists in the model output
  if (!(feature %in% rownames(summary_model_full$coefficients))) next
  
  coef_row <- summary_model_full$coefficients[feature, ]
  
  results_list_full_adeno[[feature]] <- data.frame(
    variable = feature,
    estimate = coef_row["coef"],
    std.error = coef_row["robust se"],   # IMPORTANT when using cluster()
    z = coef_row["z"],
    p.value = coef_row["Pr(>|z|)"],
    HR = exp(coef_row["coef"]),
    lower_CI = exp(coef_row["coef"] - 1.96 * coef_row["robust se"]),
    upper_CI = exp(coef_row["coef"] + 1.96 * coef_row["robust se"]),
    c_index <- summary_model_full$concordance[1]
  )
}

results_full_adeno <- do.call(rbind, results_list_full_adeno)


results_full_adeno$p_adjusted <- p.adjust(results_full_adeno$p.value, method = "bonferroni")

significant_findings_adeno <- results_full_adeno %>% filter(p_adjusted<0.05)

results_full_adeno$fdrs<-p.adjust(results_full_adeno$p.value, method="BH")
write.csv(results_full_adeno, "stats_out/cox_adeno.csv")

```
  
# Linear mixed model 
Linear mixed model was used to assess the effect of histology and stage on the features.
- includes a convergence log to assess the models performance 

```{r}
library(effectsize)
```

```{r}
# Define your response variables
response_vars <- feature_vars

# Store results
results_list  <- list()
summary_table <- data.frame()
convergence_log <- data.frame()

for (var in response_vars) {
  
  formula <- as.formula(paste(var, "~ Histology + stage_broad + (1|patientID)"))
  
  model <- tryCatch(
    lmer(formula, data = df_filtered, REML = TRUE,
         control = lmerControl(
           optimizer = "bobyqa",
           optCtrl   = list(maxfun = 2e5)
         )),
    error = function(e) {
      message("Model failed for: ", var, " — ", e$message)
      return(NULL)
    }
  )
  
  if (is.null(model)) next
  
  # Safe convergence check - the original code breaks on singular fits
  # because optinfo$conv$lme4 may be empty
  converged <- tryCatch({
    conv_messages <- model@optinfo$conv$lme4$messages
    if (is.null(conv_messages)) {
      TRUE  # no messages = clean convergence
    } else {
      !any(grepl("failed to converge", conv_messages, ignore.case = TRUE))
    }
  }, error = function(e) NA)  # NA = couldn't determine
  
  is_singular <- isSingular(model)
  
  convergence_log <- rbind(convergence_log, data.frame(
    response_var = var,
    is_singular  = is_singular,
    converged    = converged
  ))
  
  results_list[[var]] <- model
  
  tidy_output <- broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE)
  tidy_output$response_var <- var
  tidy_output$is_singular  <- is_singular
  tidy_output$converged    <- converged
  
  summary_table <- rbind(summary_table, tidy_output)
}

# FDR correction
summary_table$p_fdr <- ave(
  summary_table$p.value,
  summary_table$term,
  FUN = function(p) p.adjust(p, method = "BH")
)

# Check what you got
print(convergence_log)


write.csv(summary_table, "stats_out/lme_results_convergnce.csv", row.names = FALSE)
```

```{r}
# Get your significant variable names

significant_hits <- subset(summary_table, p_fdr < 0.05 & term != "(Intercept)")

sig_vars <- unique(significant_hits$response_var)
sig_summary <- summary_table[summary_table$response_var %in% sig_vars, ]

# For each one, print a clean summary of both fixed effects
for (var in sig_vars) {
  model <- results_list[[var]]
  cat("\n========================================\n")
  cat("Variable:", var, "\n")
  cat("========================================\n")
  
  # Fixed effects with confidence intervals
  fe <- broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE)
  fe <- subset(fe, term != "(Intercept)")
  
  print(fe[, c("term", "estimate", "conf.low", "conf.high", "p.value")])
  
  # ICC - how much variance is patient vs residual
  vc <- as.data.frame(VarCorr(model))
  icc <- vc$vcov[1] / sum(vc$vcov)
  cat("ICC (patient-level variance):", round(icc, 3), "\n")
}

```
Plot the findings 
```{r}
library(ggplot2)

# Filter to just your significant variables, exclude intercept
plot_data <- subset(summary_table, 
                    response_var %in% sig_vars & term != "(Intercept)")

ggplot(plot_data, aes(x = estimate, y = response_var, color = term)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2,
                 position = position_dodge(width = 0.5)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Estimate (with 95% CI)", y = "Response Variable",
       title = "Fixed Effects for Significant Variables",
       color = "Term") +
  theme_bw()
```

